Effectful Semantics for Substructural SSA

### 2023-11-10 pldi24/paper.tex
Cleaning, improved example programs

\usepackage{listings}
\usepackage{xcolor}
\usepackage{subfig}

@@

This ``basic blocks with arguments'' presentation of SSA has become
very popular in industry, with compiler backends such as
Cranelift~\cite{cranelift}, MLIR~\cite{mlir}, and SIL~\cite{SIL} all
using this representation. This tight connection might suggest that
the semantics of SSA representations is trivial: we can just interpret
an SSA program as a corresponding functional program, and then rewrite
the SSA programs using the equational theory of functional programs.

@@

A useful taxonomy was introduced by \citet{fuhrmann}, who classified
effects based on (1) whether or not the order of effects mattered,
(2) whether it was sound to discard an effectful operation, and (3)
whether it was sound to duplicate an effectful operation.

Suppose we substitute \texttt{print("hello")} for $x$ in the program
below:
\begin{verbatim}
let x = print("hello");            let y = print("world");
let y = print("world");   =/=>     let z = print("hello");
let z = x;                         ...
...
\end{verbatim}

In this case, substituting a term for a variable changes the order the
I/O actions are performed, which changes the program's observable
behaviour. On the other hand, suppose we have an effect which increments
a counter. Then the same transformation is sound. 
\begin{verbatim}
let x = increment(a);            let y = increment(b);
let y = increment(b);   ===>     let z = increment(a);
let z = x;                        ...
...
\end{verbatim}

F\"{u}hrmann called these commutative effects \emph{central}, which we
also call \emph{linear}.

It can also matter whether an effect is performed at all or not. For example:
\begin{verbatim}
let x = may_loop();    =/=>      return 5;
return 5 
\end{verbatim}
In this example, even though $x$ is not used, the call to \texttt{may_loop()} cannot
be erased, as the observable behaviour of the program can change if it goes into an
infinite loop.

On the other hand, if the effectful operation is merely nondeterministic, then it is
safe to discard results. 
\begin{verbatim}
let x = rand();    ===>          return 5;
return 5 
\end{verbatim}
F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use his
terminology, and also call such effects \emph{affine}.

Now consider the following example: 
\begin{verbatim}
let x = rand();    =/=>     let y = rand();
let y = x;                  let z = rand();
let z = x;                  return y == z;
return y == z
\end{verbatim}

In this case, substituting for $x$ duplicates the call to \texttt{rand()}, and can change
this from a program which always returns true to one which can return either true or false.

On the other hand, if the effect is the possibility of nontermination, then the
substitution is sound. 
\begin{verbatim}
let x = may_loop();    ===>     let y = may_loop();
let y = x;                      let z = may_loop();
let z = x;                      return y == z;
return y == z
\end{verbatim}
F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
terminology, and also call such effects \emph{relevant}.

With these distinction in place, we can \emph{define} a pure
expression as one which is all three of central, copyable, and
discardable.

In addition to working with effects, compiler intermediate
representations also have to track accesses to memory. This
information is typically gathered via ad-hoc alias analyses, but a
compositional alternative to these analyses is to track this
information via substructural type systems such as linear and
affine types, which associate aliasing and ownership information
with individual substructural values. Having aliasing information
justifies important optimisations such as loop hoisting: 
\begin{verbatim}
while(e) {              x = read(a);
  x = read(a);          while(e) { 
  body             ==>    body
}                       }
\end{verbatim}
This transformation is only sound if the body of the loop \texttt{body}
does not access \texttt{a}, which is precisely the kind of frame condition
that linear types can track.

@@

\section{Implementation}

\TODO{remove? time + space...}

>>>>

\usepackage{listings}
\usepackage{xcolor}
\usepackage{subfig}
\usepackage{fancyvrb}

@@

This "basic blocks with arguments" presentation of SSA has become very popular
in industry, with compiler backends such as Cranelift~\cite{cranelift},
MLIR~\cite{mlir}, and SIL~\cite{SIL} all using this representation. This tight
connection might suggest that the semantics of SSA representations is trivial:
we can just interpret an SSA program as a corresponding functional program, and
then rewrite the SSA programs using the equational theory of functional
programs.

@@

A useful taxonomy was introduced by \citet{fuhrmann-direct-1999}, classifying
effects based on (1) whether or not the order of effects mattered, (2) whether
it was sound to discard an effectful operation, and (3) whether it was sound to
duplicate an effectful operation. Suppose we substitute \texttt{print("hello")}
for $x$ in the program below:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = print("hello");
let y = print("world");
let z = x;   
...
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = print("world");
let x = print("hello");
...
\end{BVerbatim}
\end{equation}
In this case, substituting a term for a variable changes the order the I/O
actions are performed, which changes the program's observable behaviour. On the
other hand, suppose we have an effect which increments a counter. Then the same
transformation is sound, since \(x\) is used exactly once:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = increment(a);
let y = increment(b);
let z = x;
...
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = increment(b);
let z = increment(a);
...
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such commutative effects \emph{central}, which we also call
\emph{linear}. It can also matter whether an effect is performed at all or not.
For example, consider:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
return 5 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
In this example, even though $x$ is not used, the call to \texttt{may_loop()}
cannot be erased, as the observable behaviour of the program can change if it
goes into an infinite loop. On the other hand, if the effectful operation is
merely nondeterministic, then it is safe to discard results:
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = rand();
return 5 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
return 5
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called these operations \emph{discardable}. In this paper, we use
his terminology, and also call such effects \emph{affine}. Now consider the
following example: 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = rand();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \centernot\implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = rand();
let z = rand();
return y == z
\end{BVerbatim}
\end{equation}
In this case, substituting for $x$ duplicates the call to \texttt{rand()}, and
can change this from a program which always returns true to one which can return
either true or false. On the other hand, if the effect is the possibility of
nontermination, then the substitution is sound. 
\begin{equation}
\begin{BVerbatim}[baseline=c]
let x = may_loop();
let y = x;
let z = x;
return y == z 
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
let y = may_loop();
let z = may_loop();
return y == z
\end{BVerbatim}
\end{equation}
F\"{u}hrmann called such operations \emph{copyable}. In this paper, we use his
terminology, and also call such effects \emph{relevant}. With these distinction
in place, we can \emph{define} a pure expression as one which is all three of
central, copyable, and discardable.

In addition to working with effects, compiler intermediate representations also
have to track accesses to memory. This information is typically gathered via
ad-hoc alias analyses, but a compositional alternative to these analyses is to
track this information via substructural type systems such as linear and affine
types, which associate aliasing and ownership information with individual
substructural values. Having aliasing information justifies important
optimisations such as loop hoisting: 
\begin{equation}
\begin{BVerbatim}[baseline=c]
while (e) {
  x = read(a);
  body
}
\end{BVerbatim}
\qquad \qquad \implies \qquad \qquad
\begin{BVerbatim}[baseline=c]
x = read(a);
while(e) {
  body
}
\end{BVerbatim}
\end{equation}
This transformation is only sound if the body of the loop \texttt{body} does not
access \texttt{a}, which is precisely the kind of frame condition that linear
types can track.

@@
